Rice yield improvements by plant breeding are offset by decreases over time

### 2017-05-15 output/breed_paper.Rnw
Adds sources, re-arranged again

@@

@@

The Rice Experiment Station (RES) of the CCRRF is located northwest of
Biggs, CA (39.4648, -121.7342), and has been the central location for
the development of improved rice varieties adapted for CA since
1969. The climate at the RES is Mediterranian, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 -- 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 -- 8 farmer-managed trials per year.

@@

<<echo=FALSE>>=
rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)

m202_v_206 = round(fixef(RES_full)[2] * scl * 19, 0)
# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)

There was a positive relationship between the year of release and
yield, with an estimated \Sexpr{round(fixef(RES_full)[2] * scl, 1)}
kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
which was widely planted from the late 1980s until the early 2000s,
and M-206, currently the most widely planted CRRFS cultivar
(approximately 50\% of planted area), there is an estimated
\Sexpr{m202_v_206} kg ha\textsuperscript{-1} yield advantage of
M-206 over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).

@@

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
                           mod_data$yrs_in_trial[res])

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = fixef(RES_full)[3] * scl/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(round(fixef(RES_full)[3] * scl , 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated decrease over time once a
  cultivar is release is estimated to be
  \Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1}
  year\textsuperscript{-1}. The shaded area is the 90\% credible
  interval.}
\label{fig:RES_year}
\end{figure}

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8, 12.5),
     xlim = c(1980,2019))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * fixef(RES_full)[2] * scl)) / 1000,
       b = fixef(RES_full)[2] * scl/1000)
abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

@@

# text(1990, 8.0, bquote(.(round(fixef(RES_full)[2] * scl , 1))~kg~ha^-1~year^-1))

@@

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The solid line
  represents a simple linear regression of yield for each cultivar over
  the period which it was planted at the RES (including only that
  cultivar's data), while the dashed line is the average yield of M-202
  at its year of release (1985). Both cultivars have a similar yield at
  their respective year of release, and similar decreases over
  time. The slope for the individual regressions are comparible to the
slope estimated from all medium grain cultivars
(\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
\label{fig:compare_yrs}
\end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
@

When yields are arranged according to the years since release, there
is a small decrease in yield over time once a cultivar is released,
though the estimated effect is small compared to the variability in
the data set. This decrease is estimated to be
\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg
ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
\Sexpr{RES_CI[1]}). The oldest cultivar
in the trials is estimated to have lost
\Sexpr{round(fixef(RES_full)[3] * scl * 36 , 1)} kg
ha\textsuperscript{-1} since release.

Again comparing the yield of two prominant cultivars, M-202 and M-206,
arranged not by the year but rather the years since release, we find
that the two cultivars are compariable and that the estimated yield
advantage (\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be
explained by the differences in the amount of time that has passed
since each cultivar was released
(Fig. \ref{fig:compare_yrs} \& \ref{fig:compare_since_release})

Contrary to the estimates for decreases once a cultivar is released,
there was no increase in yield potential over time once the
differences between the years since release were accounted for (Fig. \ref{fig:release})
(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} kg ha\textsuperscript{-1}
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(RES_full)[3] * scl * 36, 1)} lbs acre\textsuperscript{-1}.

@@

\subsection{Complications}

There is strong collinearity present between the year of release and
the number of years in the trials which is complicating the
simultaneous estimation of these effects. With the current size of the
data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
the decrease over time since release and the improvement in yield over
time with each new release in the same model. This is why each effect was
estimated using a separate regression.

Increasing the size of the data to the full data set (N =
\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
and results in similar estimated effect sizes for each of these
mechanisms. However, this necessitates the inclusion of many entries
which have characteristics not similar to conventional cultivars.

@@

 We postulate that true
yield increases are only possible if the improvement in yield of each
new cultivar exceeds former's benchmark yield at the time of release,
and therefore side-by-side comparisons are potentially misleading.

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

The practical impact of these results is: Breeders need to be aware of
the potential for the yield of checks to decrease over time. Due to
this, comparisons to older checks might give a false sense of
improvement. In order to have greater improvements in yield, the
target yield improvement for a new cultivar needs to be higher to
offset the impact of yield decrease once a cultivar is
released. Breeding is essential to maintaining yield in rice. Yield is
not stable and without continuous effort, yield potential will
decrease over time.

@@

<<echo=FALSE, results=tex>>=
CIs = posterior_interval(RES_full)[1:3,] * scl
CIs[1,] = round(CIs[1,] + center, -2)
CIs = round(CIs, 1)
med_eff = fixef(RES_full) * scl
med_eff[1] = round(med_eff[1] + center, -2)
med_eff = round(med_eff, 1)

@@

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
tars <- c("M202","M206")

boxplot(yield_kg/1000 ~ id, data = mod_data[mod_data$id %in% tars,],
        ylab = expression("Grain yield ( Mg ha"^-1~")"))

@
\caption{The performance of two major cultivars at the CA Rice
Experiment Station (RES) over the period 1995 -- 2015. The newer
cultivar, M-206, appears to have a slight yield advantage (an
estimated \Sexpr{m202_v_206} kg ha\textsuperscript{-1}) over the older
cultivar, M-202, though this comparison does not account for the 10
year difference in years since release between the cultivars.}
\label{fig:compare_vars}
\end{figure}

>>>>

RES = res = mod_data$site == "RES"

@@

release_yr_coef = round(fixef(RES_full)[2] * scl, 1)
yrs_coef = round(fixef(RES_full)[3] * scl, 1)

rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
m202_v_206 = round(release_yr_coef * 19, 0)

@@

The Rice Experiment Station (RES) of the CCRRF is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted for CA since
1969. The climate at the RES is Mediterranian, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 -- 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 -- 8 farmer-managed trials per year.

@@

@@

plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
                           mod_data$yrs_in_trial[res])

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = yrs_coef/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated decrease over time once a
  cultivar is release is estimated to be
  \Sexpr{yrs_coef} kg ha\textsuperscript{-1}
  year\textsuperscript{-1}. The shaded area is the 90\% credible
  interval.}
\label{fig:RES_year}
\end{figure}

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8, 12.5),
     xlim = c(1980,2019))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000)
abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

@@

# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))

@@

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The solid line
  represents a simple linear regression of yield for each cultivar over
  the period which it was planted at the RES (including only that
  cultivar's data), while the dashed line is the average yield of M-202
  at its year of release (1985). Both cultivars have a similar yield at
  their respective year of release, and similar decreases over
  time. The slope for the individual regressions are comparible to the
slope estimated from all medium grain cultivars
(\Sexpr{yrs_coef} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
\label{fig:compare_yrs}
\end{figure}

<<echo=FALSE>>=
RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
@

When yields are compared accounting for differences in the years since
release at the time of comparision, there is a small decrease in yield
over time once a cultivar is released
(\Sexpr{yrs_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1}, 90\%
CI: \Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in
the trials is estimated to have lost \Sexpr{round(yrs_coef*36,1)}
kg ha\textsuperscript{-1} since release. Again
comparing the yield of two prominant cultivars, M-202 and M-206, the
two cultivars are experiencing a similar trend over time to eachother
(Fig. \ref{fig:compare_yrs}) and the estimate from all released medium
grains. The estimated yield advantage of the newer cultivar (M-206)
(\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be explained by the
differences in the amount of time that has passed since each cultivar
was released (Fig. \ref{fig:compare_yrs} \&
\ref{fig:compare_since_release})

<<echo=FALSE>>=
# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)

Contrary to the estimates for decreases once a cultivar is released,
there was no increase in yield potential over time once the
differences between the years since release were accounted for
(Fig. \ref{fig:release}) (\Sexpr{release_yr_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1}). This suggests
that the overall progress on yield over the period from 1980 -- 2016
is approximately \Sexpr{round(release_yr_coef * 36, 0)} kg
ha\textsuperscript{-1}.

When differences between the number of years since each cultivar's
release are taken into account, the estimated increase in grain yield
over time is estimated to be \Sexpr{release_yr_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]}). Therefore, the evidence is
consistent with no increase in yield potential over the study
period. If the differences in years since release of each cultivar is
ignored, there appears to be a small increase over time (approximately
\Sexpr{round(coef(lm(yield_kg ~ release_yr, mod_data[res,])), 1)})
(Fig. \ref{fig:release}), much less than previous estimates of
\Sexpr{54 * 1.12} kg ha\textsuperscript{-1} year\textsuperscript{-1}
(CITE).  Comparing the two dominate cultivars M-202, which was widely
planted from the late 1980s until the early 2000s, and M-206,
currently the most widely planted CRRFS cultivar (approximately 50\%
of planted area), there is an estimated \Sexpr{m202_v_206} kg
ha\textsuperscript{-1} yield advantage of M-206 over M-202 (Figs
\ref{fig:release} \& \ref{fig:compare_vars}), though both cultivars
yielded similarly at their respective year of release
(Fig. \ref{fig:compare_yrs}).

@@

\subsection{Improving yield potential over time}

Overall, we found evidence consistent with a no improvements in yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])}. Rather, it seems possible that the
breeding program has been maintaining yield potential in the face of
changing biotic and abiotic pressures, which are expected to decrease
yield over time \citep{Peng2010}. This highlights both the need
for maintanence breeding and the difficulty of increasing genetic
yield potential as yield approaches the physiological limit. Without
maintanence breeding continuously selecting cultivars well adapted to
current conditions, CA rice production systems are predicted to have
lost an estimated \Sexpr{round(yrs_coef * 31, 0)} kg
ha\textsuperscript{-1} grain yield over the study period.

However, the goal of a breeding program is to not only recover the
benchmark yield of the old standard cultivar but to surpass it. Given
the estimated yield decreases over time, to hit an ambitious target of
500 kg ha\textsuperscript{-1} (approximately 5\%) increase in yield
per release (with roughly 5 years between releases), a new cultivar
needs to yield \Sexpr{round(yrs_coef * -5, -1)} more
in side-by-side comparison on top of the 500 kg ha\textsuperscript{-1}
target improvement. Often side-by-side yield improvements (not
considering the difference in years since release) are closer to 2 --
3\% per release, in which case no absolute improvements might be made
beyond maintenance. The evidence here is consistent with the latter
situation occuring in CA, which suggests that the genetic yield
potential has not increased substantially in the last 30+ years (Fig. \ref{fig:release}).

There has been some yield improvements over the last 30 years in the
CA state average yield (Fig. \ref{fig:state_yield}), but the evidence
here suggests mechanisms other than increased genetic yield potential
are responsible (Fig. \ref{fig:compare_yrs}). Improvement in agronomic
management, including more efficient nutrient management
\citep{Lundy2012, Lundy2015, linquist2009}, weed management
\citep{Saito2008, Pittelkow2012, linquist2008}, etc. may all be
contributing to further increases. However, the gains here are less
than those experienced during the ``Green revolution'', and by their
nature cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system, state average yields
(approximately 10 Mg ha\textsuperscript{-1} at the time of writing;
Fig. \ref{fig:state_yield}) may be approaching the agronomic limit.

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other improvements due to plant
improvement, including qualtity \citep{mckenzie1993} and milling yield
\citep{mckenzie1994} have been made over the study period and are no
less important. However, planning to meet future demand necessitates
increases in grain yield \citep{Godfray2010, Foley2011a,
Mueller2012}. The evidence here suggests that to accomplish true
increases over time, the improvement target needs take decreases in
the current cultivars' yield over time into account. While side-by-side
comparisions will continue to be useful, comparisons across years to
each cultivar's benchmark yield at the time of release should provide
information on the absolute improvements being made in yield potential over time.

\begin{figure}[ht]
<<fig = TRUE, echo = FALSE>>=

plot(1, type = "n",
     xaxt = "n",
     xlab = "Time",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8.5,11.5),
     xlim = c(0,40),
     xaxs = "i")

arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
text(25, 10.75, "(3) target above maintenance", cex = 0.8,
     srt = 15)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
abline(h = 10, lty = 3, lwd = 2)
text(25, 10.1, "(1) current situation", cex = 0.8)
abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
text(25, 9.25, "(2) no maintenance breeding", cex = 0.8,
     srt = -20)

@ 
\caption{A conceptual diagram highlighting three scenarios:
  (1) maintenance breeding, where losses over time are offset by
  improvements via plant improvement, (2) no maintenance breeding,
  where grain yield slowly decreases over time, and (3) absolute
  improvement in grain yield, where gains in grain yield exceed the
  losses over time.}
\label{fig:theory}
\end{figure}

@@

Improvement in grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. To achieve these
improvements, we show here that (1) continuous maintenance via plant
improvement is required, without which yields would decrease over
time, (2) the challenge of improving genetic yield potential is
substantial, since absolute gains require improvement in yield in
excess of the maintenance threshold, and (3) increasing yield through
improved management (agronomy) cannot exceed the ceiling of genetic
yield potential. Taken together, these conclusions underscore both the
necessity and substantial challenges facing rice breeding. We suggest
here that historical benchmarks be used in to evaluate promising rice
cultivars rather than side-by-side comparisions to maintain
perspective of the absolute gains being made.

@@

<<echo=FALSE, results=tex>>=

CIs = posterior_interval(RES_full)[1:3,] * scl
CIs[1,] = round(CIs[1,] + center, -2)
CIs = round(CIs, 1)
med_eff = fixef(RES_full) * scl
med_eff[1] = round(med_eff[1] + center, -2)
med_eff = round(med_eff, 1)

@@

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
tars <- c("M202","M206")
comp_yrs = 2002:2005

boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
                                            (mod_data$year %in% comp_yrs) &
                                            mod_data$site == "RES"  ,],
        ylab = expression("Grain yield ( Mg ha"^-1~")"))

@
\caption{An example of a typical comparison between a older (M-202)
  and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
  Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
  \Sexpr{max(comp_yrs)}, approximately the comparision made to
  determine if a new cultivar performs sufficiently better. The newer
  cultivar, M-206, appears to have a yield advantage over the older
  cultivar, M-202, though this comparison does not account for the 10
  year difference in years since release between the cultivars.}
\label{fig:compare_vars}
\end{figure}
